---
title: "Report Part 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Question 1
To approach this, we have a tsv file, first column and row are the names, so we must account for that using header and row.names. The tsv is read and assigned to a variable for easy usage.
```{r}
URL="https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"  
gene_expression_raw <- read.table(URL,stringsAsFactors = FALSE, header=TRUE, row.names=1)  #Reading the URL and formatting it correctly
head(gene_expression_raw) #Show the first 6 genes in the tbale
```

### Question 2
To approach this, we create a new column called "Mean_Expression" in "gene_expression" and  assign the mean values to this column. 
```{r}
gene_expression <- gene_expression_raw #Create a new data set to modify, leave the original
gene_expression$Mean_Expression <- rowMeans(gene_expression_raw, na.rm = TRUE) #Finding the means of the rows, remove the empty values
head(gene_expression)
```


### Question 3

To approach this, we used the order function, and use - prefix to indicate decreasing order of Mean_Expression, then took the first 10 using head. The 10 genes are summarised in the table below:
```{r}
sorted_gene_expression <- gene_expression[order(-gene_expression$Mean_Expression), ] #Using the order function, sort the Mean expression in decreasing order
head(sorted_gene_expression, 10) #Show the first 10
```



### Question 4
We first filtered genes with a mean expression less than 10, then summed the number of TRUE values.
```{r}
sum(sorted_gene_expression$Mean_Expression <10) #Use sum function which counts the number of TRUE expressions, each row with Mean_Expression less than 10 will be TRUE
```
We found that A total of 35,988 genes have a mean expression less than 10.





### Question 5


### Question 6


### Question 7


### Question 8



### Question 9


### Question 10



