---
title: "R Notebook"
output: html_notebook
---

Load the necessary libraries:

```{r}
suppressPackageStartupMessages({
  library("seqinr") # is a package designed to process and analyse sequence data.
  library("R.utils") # general utilities like zip and unzip
  library("ggplot2")
})
```
1. Read in the file, making the gene identifiers the row names. Show a table of values for the first six genes

```{r}
URL="https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/gene_expression.tsv"
gene_expression_raw <- read.table(URL,stringsAsFactors = FALSE, header=TRUE, row.names=1)
head(gene_expression_raw)
```

2. Make a new column which is the mean of the other columns. Show a table of values for the first six genes.

```{r}
gene_expression <- gene_expression_raw
gene_expression$Mean_Expression <- rowMeans(gene_expression_raw, na.rm = TRUE)
head(gene_expression)
```

3.  List the 10 genes with the highest mean expression

```{r}
sorted_gene_expression <- gene_expression[order(-gene_expression$Mean_Expression), ]
head(sorted_gene_expression,10) 
```


4. Determine the number of genes with a mean <10
```{r}
sum(sorted_gene_expression$Mean_Expression <10) 
```

5. Make a histogram plot of the mean values

```{r}
# histogram
hist(gene_expression$Mean_Expression, xlab = "Mean Expression", main = "Histogram of Gene Expression Means")
```

6. Import this csv file into an R object. What are the column names?

```{r}
URL= "https://raw.githubusercontent.com/ghazkha/Assessment4/refs/heads/main/growth_data.csv"
growth_data_raw <- read.csv(URL)
colnames(growth_data_raw)
```





